---
title: "Blackjack"
author: "Andrea Roveroni"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# The deck
A deck consists of 52 cards of 4 different suits, which are:
- 1 * 4 aces
- 9 * 4 numeric cards, from 2 to 10
- 3 * 4 face cards, Jack, Queen and King

Aces can either value 1 or 11 (whichever is favorable), numeric cards value their number, face cards all value 10

# Blackjack
A blackjack happens when the first two cards in your hand score 21. This means you need to have an ace and either a figure or a 10

So, to have a Blackjack given a full deck of 52 cards, you need to pick:
- 1 ace from the 4 available
- 1 10 from the 16 available

This leads us to the probability:
$$
Blackjack = \frac {\binom {4} {1} * \binom {16} {1}} {\binom {52} {2}} = `r (choose(4, 1) * choose(16, 1)) / choose(52, 2)`
$$

Since the game is generally played with 6-8 decks, we can generalize this formula for `n` decks:
$$
Blackjack = \frac {\binom {4n} {1} * \binom {16n} {1}} {\binom {52n} {2}} = \frac {128n^2} {52n * 52(n-1)}
$$

And we can plot it:
```{r}
x = 1:10
y = (choose(4*x, 1) * choose(16*x, 1)) / choose(52*x, 2)

plot(x, y)
```

We can observe that the probability of a Blackjack decreases as the number of decks increases.
However, note that it will never reach zero.
**TODO: Show limit `x -> inf`**


# Theorem 8.1 (Infinite Deck Assumption)

The infinite deck assumption in Blackjack fixes the probability of getting any non-ten-valued card as `4/52` and getting a ten-valued card as `16/52`. This assumption introduces very little error in calculation, and can be used to determine strategies for Blackjack games with a large number of decks.


# Decision making

## Hitting or standing

Suppose we have in our hand an 8 and a 10, and that the dealer has a Q.

If we hit, meaning if we take another card, we will bust for any card > 3. So, to not bust, we must pick an ace, a 2 or a 3. Using the Infinite Deck Assumption theorem, we can thus calculate the probability of not busting after we decided to pick a card:
$$
notBustAfterHit = \frac {4} {52} * 3 = `r notBustAfterHit = 4/52 * 3; notBustAfterHit`
$$

And the probability of busting is the opposite:
$$
bustAfterHit = \bar {notBustAfterHit} = `r bustAfterHit = 1 - notBustAfterHit; bustAfterHit`
$$

We want now calculate the probability of losing if we stand, so we can compare both options. We lose if the dealer scores > 18 (if the dealer scores 18 like us, the round ends with a tie) and does not bust.
** See page 227**

We end up with the probability of the dealer winning as `0.488119`

So:
$$
loseAfterStanding = `r loseAfterStanding = 0.488119; loseAfterStanding`
$$

Comparing both options, we have a probability of losing after hitting of `r bustAfterHit` and a probability of losing after standing of `r loseAfterStanding`.

This means that we are surely better standing.

Note that we can also have a tie if the dealer scores 18, so the probability of winning is not the opposite of losing after standing.

The probability that the dealer scores 19 can be calculated similarly as before:
$$
tieAfterStanding = `r tieAfterStanding = 0.111424; tieAfterStanding`
$$

We can calculate the expected winnings by standing. Assuming `1.00 $` bet:
$$
expectedWinningsForStand = (1 - loseAfterStanding - tieAfterStanding) * 1.00 + tieAfterStanding * 0.00 + loseAfterStanding * (-1.00) = `r (1 - loseAfterStanding - tieAfterStanding) * 1.00 + loseAfterStanding * (-1.00)`
$$

This gives the advantage to the dealer, but note that the dealer would have an even greater advantage if we decided to hit.

# Probability for the dealer of a specific score

**First, always initialize the following**

```{r}
# Current face up card of the dealer. Additionally, we assume the dealer has already checked for blackjack on the hole card
dealer_up_card = 10

# The expected sum of the dealer's cards for which to calculate the probability
target_dealer_value = 19

# Total number of decks in the game
# Assumption: The current face up cards (both dealer's and players') currently do neither affect the available cards, nor the probability for each card, that we assume as stated before, 16/52 for each ten-valued card, 4/52 for each non-ten-valued card
num_decks = 1

# If true, the dealer stands if it has a soft 17
soft_stands = FALSE
```

So the problem becomes:

**Given the dealer has a $`r dealer_up_card`$ as the up card, and we play with $`r num_decks`$ deck(s), what is the probability of the dealer to score a total of $`r target_dealer_value`$?**


```{r}
# This function computes all possible ordered card extractions that give the expected value
# Note that cards are ordered by their value, so here we're missing possible permutations
# Additionally, no constraints are checked, like available cards, blackjack, dealer stand
find_combinations <- function(values, target) {
  find_recursive <- function(subset, remaining_values, target) {
    # If we reached the target, end the branch here
    if (sum(subset) == target) {
      return(list(subset))
    }
    # If we're above the target, the entire branch is not valid
    if (sum(subset) > target) {
      return(list())
    }
    
    solutions = list()
    
    # Loop for each remaining value
    for (i in seq_along(remaining_values)) {
      new_subset = c(subset, remaining_values[i])
      # Try a branch with this value
      solutions = c(solutions, find_recursive(new_subset, remaining_values[i:length(remaining_values)], target))
    }
    
    return(solutions)
  }
  
  return(find_recursive(c(), values, target))
}

# Helper function that, given some extraction paths (branches) and the amount of each card, returns only the paths that do not exceed the available amount for each card 
remove_paths_exceeding_cards <- function(paths, available_cards) {
  return(Filter(function(path) {
      used_cards = c(rep(0, length(available_cards)))
      
      for(i in path){
          # Ace and 1 are the same card
          index = ifelse(i == 11, 1, i)
          
          used_cards[[index]] = used_cards[[index]] + 1
          
          # If the current card has been used more than the max number of times it is available, then the path is not valid
          if(used_cards[[index]] > available_cards[[index]]){
            return(FALSE)
          }
      }
      
      return(TRUE)
    }, paths))
}
```

**TODO: Not exactly, redo with new logic:**
If the dealer has, for example, a Q, to make 19 he has the following possibilities:

```{r}
# First, calculate all possible combinations that give a certain target
possible_cards = 1:11
paths = find_combinations(possible_cards, target_dealer_value - dealer_up_card)


# Then, depending on the number of decks, filter out combinations that exceed the maximum number of each card

# For each deck, we have 4 cards from 1 to 9, and 16 ten-valued cards. Ace counts as value one
available_cards = c(rep(4, 9), 16)* num_decks

valid_paths = remove_paths_exceeding_cards(paths, available_cards)
```

```{r}
# Given a path, returns the corresponding probability for it to occur
calculate_path_probability <- function(path){
  prob = 1
  for(card in path){
    # We said that for any 10-valued card we have a 16/52 probability
    if(card == 10){
      prob = prob * (16/52)
    }
    # And a 4/52 probability for any non-10-valued card
    else{
      prob = prob * (4 / 52)
    }
  }
    
  return(prob)
}

# Given a path, returns the total numbers of its anagrams (without repetition)
count_path_anagrams <- function(path){
  element_occurrences = as.vector(table(path))
  
  return(factorial(length(path)) / prod(factorial(element_occurrences)))
}
```


The following is the code, wrapped in a function so to be reused later
```{r}
calc_probability_exact_dealer_score <- function(dealer_up_card, target_dealer_value, num_decks, cache){
  # Note: the dealer always hits on 17, no matter soft 17
  cache_key = paste(dealer_up_card, target_dealer_value, num_decks, sep = "_")
  if(exists(cache_key, envir = cache)){
    return(get(cache_key, envir = cache))
  }
  
  # If the dealer already has an higher value then the up card, he will surely make a higher score
  if(target_dealer_value < dealer_up_card){
    assign(cache_key, 0, envir = cache)
    return(0)
  }
  
  # Instead, if the dealer already has the up card as the target value,
  # the probability is 1 since he just does not need to pick any card
  if(target_dealer_value == dealer_up_card){
    assign(cache_key, 1, envir = cache)
    return(1)
  }
  
  # For target values < 17, we do not need to consider the rule of standing when scoring 17 or above
  if(target_dealer_value < 17){
    # Calculate all possible combinations for the target value
    # We remove the possibility to an ace to be considered 1. We'll calculate it later
    possible_cards = 2:11
    paths = find_combinations(possible_cards, target_dealer_value - dealer_up_card)
    
    # Then, filter out combinations that exceed the available cards
    available_cards = c(rep(4, 9), 16)* num_decks
    valid_paths = remove_paths_exceeding_cards(paths, available_cards)
    
    total_probability = 0
    
    for(path in valid_paths){
      path_probability = calculate_path_probability(path)
      
      path_anagrams_count = count_path_anagrams(path)
      
      # We should prevent blackjacks
      path_anagrams_not_blackjack_count = path_anagrams_count
      if(dealer_up_card == 10 && any(path %in% c(1, 11))){
        first_occ <- which(path %in% c(1, 11))[1]
        
        path_anagrams_not_blackjack_count = path_anagrams_not_blackjack_count - count_path_anagrams(path[-first_occ])
      }
      if(any(dealer_up_card %in% c(1, 11)) && any(path %in% 10)){
        first_occ <- which(path %in% c(10))[1]
        
        path_anagrams_not_blackjack_count = path_anagrams_not_blackjack_count - count_path_anagrams(path[-first_occ])
      }
      
      # print(cat("Given up card ", dealer_up_card, " and target ", target_dealer_value, "and path ", toString(path)," anagram counts are ", path_anagrams_not_blackjack_count))
      
      total_probability = total_probability + path_probability * path_anagrams_not_blackjack_count
    }
    
    # Now we count the probability of an ace (to be picked as the last card) to be counted as one
    # This is only possible if the ace would make the dealer reach a score higher than 21
    if((target_dealer_value - 1 + 11) > 21){
      probability_minus_one = calc_probability_exact_dealer_score(dealer_up_card, target_dealer_value - 1, num_decks, cache)
      
      # print(cat("For target value ", target_dealer_value, " we could also have a 1-ace with following prob for -1: ", probability_minus_one))
      
      total_probability = total_probability + probability_minus_one * 4 / 52
    }
  
    assign(cache_key, total_probability, envir = cache)
    return(total_probability)
  }
  else{
    # If instead we hit 17 or above, we need to split the computation:
    # The dealer should reach a score lower than 16, and then hit a single card without busting
    safe_scores = Filter(function(score){
      return(score > 0 && score < 17)
    }, target_dealer_value - 1:11)
    total_probability = 0
    
    for(score in safe_scores){
        card_difference = target_dealer_value - score
        
        # We also need to ignore the blackjack combination
        is_blackjack = (dealer_up_card == 10 && dealer_up_card == score && card_difference == 11) ||
          (dealer_up_card %in% c(1, 11) && dealer_up_card == score && card_difference == 10)
        
        if(!is_blackjack){
          # print(cat("For standing at ", target_dealer_value, " we can reach ", score, " and then pick ", card_difference))
          
          card_probability = if(card_difference == 10) {16/52} else {4/52}
          
          # Here we calculate the probability of obtaining a certain score
          intermediate_score_prob = calc_probability_exact_dealer_score(dealer_up_card, score, num_decks, cache)
          # print(cat("The above has probability ", intermediate_score_prob, " * card_probability"))
          
          
          total_probability = total_probability + intermediate_score_prob * card_probability
        }
        
    }
    
    assign(cache_key, total_probability, envir = cache)
    return(total_probability)
  }
}
```


# Probability of the player to bust

This is a simpler question, since we do not need to calculate all possible combinations, since we're only interested in busting with a single card.

Thus, given the current player score, we can calculate all possible card values that will exceed 21

**First, always initialize the following**

```{r}
# Sum of current cards o the player
player_score = 18
```

```{r}
calc_probability_player_busting <- function(player_score){
  busting_probability = 0

  # Note that if we pick an ace, we only consider it as a 1
  for(i in 1:10){
    if(player_score + i > 21){
      # We busted with a card i-valued
      # We use the infinite deck theorem to calculate the probability of this i-valued card
      if(i == 10){
        busting_probability = busting_probability + 16/52
      }
      else{
        busting_probability = busting_probability + 4/52
      }
    }
  }
  
  return(busting_probability)
}

busting_probability = calc_probability_player_busting(player_score)
```

**So, given the player score of $`r player_score`$, the probability of busting is $`r busting_probability`$**

# Comparing hitting and standing

We are now able to compute the probability for each of the following cases:

1. If we stand:
    
    1.1 Probability of the dealer to score higher than us (player loses)
    
    1.2 Probability of dealer scoring exactly like us (tie)
    
    1.3 Probability of the dealer to score less than us or exceeds 21 (player wins)
  
2. If we hit:
    
    2.1 Probability of busting (player loses)
    
## 1.1: Lose after standing

The probability of losing after standing is the sum of the probabilities in which the dealer scores higher than the player:

```{r}
calc_probability_lose_after_standing <- function(player_score, dealer_up_card, num_decks, cache){
  prob = 0
  
  for(i in (player_score + 1):21){
    if(i >= 17){
      prob = prob + calc_probability_exact_dealer_score(dealer_up_card, i, num_decks, cache)
    }
  }
  
  return(prob)
}

prob_lose_after_standing = calc_probability_lose_after_standing(player_score, dealer_up_card, num_decks, new.env())
prob_lose_after_standing
```

## 1.2 Tie after standing

The probability of tie after standing is the probability the dealer scores exactly like the player:

```{r}
calc_probability_tie_after_standing <- function(player_score, dealer_up_card, num_decks, cache){
  if(player_score < 17){
    return(0)
  }
  
  prob = calc_probability_exact_dealer_score(dealer_up_card, player_score, num_decks, cache)
  
  return(prob)
}

prob_tie_after_standing = calc_probability_tie_after_standing(player_score, dealer_up_card, num_decks, new.env())
prob_tie_after_standing
```


## 1.3 Win after standing

Since this is the event in which the player neither loses nor ties, it can be calculated as:

```{r}
prob_win_after_standind = 1 - prob_lose_after_standing - prob_tie_after_standing
prob_win_after_standind
```


## 2.1 Losing after hitting

This is the event in which the player exceeds 21 after deciding to hit:
```{r}
prob_lose_after_hit = calc_probability_player_busting(player_score)
prob_lose_after_hit
```


# Deciding which move is more convenient

To summarise, we had the following situation:

- The dealer had a $`r dealer_up_card`$-valued card as the face-up card. We're also assuming the rule that the dealer peeks the hole card to see if it's a blackjack

- The player has some cards for a total of $`r player_score`$

- The rule "dealer stands if has soft 17" is $`r soft_stands`$

- We're playing with $`r num_decks`$ deck(s)

If the player stands, the probability of losing is $`r prob_lose_after_standing`$. If the player hits, the probability of losing is $`r prob_lose_after_hit`$.

The right choice to play is the one that gives the lower chances of losing, so in this case, the player should $`r if (prob_lose_after_standing < prob_lose_after_hit){"stand"} else {"hit"}`$.


# Expanding for all possibilities

We are now able to automatically compute the best strategy for each pair of $(playerScore, dealerUpCard)$:
```{r}
# We can safely ignore a score of 21 for the player
possible_player_scores = 1:20

possible_dealer_face_up_card = 1:11 # 7:11 works on single core

num_decks = 1
soft_stands = FALSE
```

The following computes the decision matrix:

```{r}
cache = new.env()
decision_matrix = sapply(possible_dealer_face_up_card, function(dealer_up_card){
  return(sapply(possible_player_scores, function(player_score){
    prob_lose_after_standing = calc_probability_lose_after_standing(player_score, dealer_up_card, num_decks, cache)
  
    prob_lose_after_hit = calc_probability_player_busting(player_score)
    
    return(prob_lose_after_standing - prob_lose_after_hit)
    
    if(prob_lose_after_standing < prob_lose_after_hit){
      return("S")
    }
    else{
      return("H")
    }
  }))
})

decision_matrix
```

We can thus print the matrix:
```{r}
if (!require(ggplot2)) {
  install.packages("ggplot2")
}
if (!require(reshape2)) {
  install.packages("reshape2")
}
library(ggplot2)
library(reshape2)

df <- melt(sign(decision_matrix))
colnames(df) <- c("Y", "X", "Value")

# 
ggplot(df, aes(x = X, y = Y, fill = Value)) +
  geom_tile(color = "grey") +                       # Crea le celle con bordi bianchi
  scale_fill_gradientn(colors = c("white", "black"), # Scala colore da nero a bianco
                       values = c(0, 1),             # Mappa 0 (nero) e valori positivi (bianco)
                       limits = c(-1, 1)) +          # Imposta i limiti della scala dei colori
  #scale_y_reverse() +                                # Inverte l'asse Y per un aspetto simile a una matrice
  scale_x_continuous(breaks = 1:ncol(decision_matrix), labels = possible_dealer_face_up_card) +  # Mostra tutte le colonne
  scale_y_continuous(breaks = 1:nrow(decision_matrix), labels = possible_player_scores) +  # Mostra tutte le righe
  theme_minimal() +
  labs(x = "Dealer's up card", y = "Hard total", fill = "Hit if > 0, else Stand") + # Aggiunge etichette
  theme(panel.grid = element_blank())                # Rimuove la griglia di sfondo
```

# TODO: Algoritmo calcolo permutazioni
Dato un valore target X:
- Enumerare le partizioni di x (find_combinations)
- Per ogni partizione, calcolare "gli anagrammi"con formula:
$$
anagrammi = \frac {l!} {k_1! * k_2! * ... * k_n!}
$$
- Per ogni partizione, la probabilità dei suoi anagrammi è data da:
prob_partizione * num_anagrammi

- Dove la probabilità della partizione è data:
prob_partizione: (16/52) ^ (num_carte_10) * (4/52) ^ (num_carte_not_10)


